#Alison's Project 1 Report

```{r TotalSteps}
        library(ggplot2)
        library(dplyr)
        library(lattice)
        
        ##read the data into a dataframe dat
        dat <- read.csv("activity.csv", stringsAsFactors = FALSE)
        
        ##convert the date variable into a date type
        dat$date <- as.Date(dat$date, "%Y-%m-%d")
        
        ##calculate the total number of steps taken
        totsteps <- sum(dat$steps, na.rm = TRUE)
```

There were **`r totsteps`** total steps taken in the dataset.

```{r Histogram}

        ##calculate the total steps per day
        
        s <- split(dat, dat$date)
        stepsday <- sapply(s, function(x) sum(x[,"steps"]))
        
        ##use this data to plot a histogram of total steps per day
        qplot(stepsday, xlab="Total Steps Per Day", main="Total Steps Per Day", binwidth=1000)
```

```{r meanmedian}
        ##calculate the mean and median of the total steps per day
        tsmean <- as.integer(mean(stepsday, na.rm=TRUE))
        tsmedian <- median(stepsday, na.rm=TRUE)
```

The mean of the total steps per day is: **`r tsmean`**
The median of the total steps per day is: **`r tsmedian`**

```{r linegraph}
        ##calculate the mean steps by interval over all time
        mytab <- summarise(group_by(dat,interval), mean(steps, na.rm=TRUE))
        names(mytab) <- c("interval", "mean.steps")
        ##make the plot of Average number of steps by interval
        plot(mytab$interval, mytab$mean.steps, type="l", ylab = "Mean number of Steps", 
             xlab="Interval", main="Average number of steps per day by Interval")
```

```{r max.interval}
        ##find the maximum value of the average number of steps
        mymax <- max(mytab$mean.steps)
        ##find the row containing this max
        myrow <- filter(mytab, mean.steps == mymax)
        ##hence find the interval containing the maximum number of steps
        myint <- myrow[,1]
```

The interval containing the maximum number of steps is: **`r myint`**

##Missing Values

```{r NAs}
        ##calculate the number of rows with some NAs in them
        badrows <- nrow(dat) - sum(complete.cases(dat))
```

There are **`r badrows`** rows containing NAs

###Strategy for Addressing Missing Values

My strategy for addressing missing values is to populate the NAs with the mean for that interval.

```{r replace.missing}
        ##replace the missing values in steps variable with the mean number of steps for that interval
        ##we will use mytab from above to lookup the mean values per interval
        dat2 <- dat
                for (i in seq_len(nrow(dat))) {
                        if (is.na(dat[i,"steps"])) { 
                                dat2[i,"steps"] <- mytab[(mytab$interval==dat[i,"interval"]),2]
                        }     
                }
             
```

Plot another histogram as per the start of the report but now using the dataset with the missing values imputed with interval means.

```{r new.histogram}
        ##calculate the total steps per day
        
        s2 <- split(dat2, dat2$date)
        stepsday2 <- sapply(s2, function(z) sum(z[,"steps"]))
        
        ##use this data to plot a histogram of total steps per day
        qplot(stepsday2, xlab="Total Steps Per Day", main="Total Steps Per Day", binwidth=1000) 
        
        ##calculate the mean and median of the total steps per day
        tsmean2 <- as.integer(mean(stepsday2, na.rm=TRUE))
        tsmedian2 <- as.integer(median(stepsday2, na.rm=TRUE))
        
```        

By comparing the two histograms we can see that the total steps per day increases at around the 1200 steps mark in the imputed dataset.

The new mean is `r tsmean2` and is the same as the new median `r tsmedian2`. Both of these are the same as the original mean. This is not surprising, perhaps, as we have used mean to populate missing values. The new median is a litte higher than the original median.

##Weekdays versus weekends

```{r weekdays}
        ##decide the day of the week each date is
        myday <- weekdays(dat2$date)
        ##replace the days with either weekday or weekend as appropriate
        for (i in 1:length(myday)) {
                if (myday[i] == "Saturday" | myday[i] == "Sunday") {
                        myday[i] <- "weekend"       
                }
                else {
                        myday[i] <- "weekday"
                }
                i <- i + 1
        }
        
##now bind this vector as a column to dat2
        dat2 <- cbind(dat2, myday)
        
        ##make a dataframe containing the mean steps per interval for all days dependent on weekday
        ##and weekend
        mytab3 <- summarise(group_by(dat2,interval, myday), mean(steps, na.rm=TRUE))
        names(mytab3) <- c("interval", "myday", "meansteps")
        ##build the lattice plot to show weekday versus weekend
        xyplot(meansteps ~ interval | myday, data = mytab3, type = "l", 
               ylab = "Number of Steps", layout = c(1,2))
        
```        


